DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-06-28 manuscript/manuscript-1st-submission.tex
add updated files for review submission

\linenumbers
\usepackage{csquotes}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=english]{babel}
\else
  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\fi

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain, 2018),
geiger (Pennell et al., 2014),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

@@

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string. Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.

@@

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

@@

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).

@@

Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

@@

\textbf{\emph{Summarizing search results.--}}
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).

@@

\textbf{\emph{Creating a query.--}}
To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group. Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.

@@

\textbf{\emph{Summarizing search results.--}}
The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).

@@

We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).

@@

In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al. (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
Alström, P., Hooper, D. M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T. D. (2014). Discovery of a relict lineage and monotypic family of passerine birds. \emph{Biology Letters}, \emph{10}(3), 20131067.

\leavevmode\hypertarget{ref-ane2009groves}{}%
Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.

\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
Antonelli, A., Hettling, H., Condamine, F. L., Vos, K., Nilsson, R. H., Sanderson, M. J., \ldots{} Vos, R. A. (2017). Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa. \emph{Systematic Biology}, \emph{66}(2), 153--166. \url{https://doi.org/10.1093/sysbio/syw066}

\leavevmode\hypertarget{ref-archie1986newick}{}%
Archie, J., Day, W. H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F. J., \& Swofford, D. (1986). The Newick tree format. Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}

\leavevmode\hypertarget{ref-avibase-emberiza}{}%
Avibase. (2022). Yellow-throated Bunting. \emph{Avibase - the World Bird Database}, (Online Resource). Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}

\leavevmode\hypertarget{ref-Bapst2012a}{}%
Bapst, D. W. (2012). Paleotree: An R package for paleontological and phylogenetic analyses of evolution. \emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807. \url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}

\leavevmode\hypertarget{ref-barba2018constraining}{}%
Barba-Montoya, J., Reis, M. dos, Schneider, H., Donoghue, P. C., \& Yang, Z. (2018). Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution. \emph{New Phytologist}, \emph{218}(2), 819--834.

\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.

\leavevmode\hypertarget{ref-barker2013going}{}%
Barker, F. K., Burns, K. J., Klicka, J., Lanyon, S. M., \& Lovette, I. J. (2013). Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds. \emph{Systematic Biology}, \emph{62}(2), 298--320.

\leavevmode\hypertarget{ref-barker2015new}{}%
Barker, F. K., Burns, K. J., Klicka, J., Lanyon, S. M., \& Lovette, I. J. (2015). New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. \emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.

\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
Barker, F. K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J. (2004). Phylogeny and diversification of the largest avian radiation. \emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.

\leavevmode\hypertarget{ref-beresford2005african}{}%
Beresford, P., Barker, F., Ryan, P., \& Crowe, T. (2005). African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''. \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.

\leavevmode\hypertarget{ref-Boyle2013}{}%
Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J. A., Mozzherin, D., Rees, T., \ldots{} Enquist, B. J. (2013). The taxonomic name resolution service: An online tool for automated standardization of plant names. \emph{BMC Bioinformatics}, \emph{14}(1). \url{https://doi.org/10.1186/1471-2105-14-16}

\leavevmode\hypertarget{ref-Britton2007}{}%
Britton, T., Anderson, C. L., Jacquet, D., Lundqvist, S., \& Bremer, K. (2007). Estimating Divergence Times in Large Phylogenetic Trees. \emph{Systematic Biology}, \emph{56}(788777878), 741--752. \url{https://doi.org/10.1080/10635150701613783}

\leavevmode\hypertarget{ref-bryson2014diversification}{}%
Bryson Jr, R. W., Chaves, J., Smith, B. T., Miller, M. J., Winker, K., Pérez-Emán, J. L., \& Klicka, J. (2014). Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae). \emph{Journal of Biogeography}, \emph{41}(3), 587--599.

\leavevmode\hypertarget{ref-burleigh2015building}{}%
Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.

\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
Burns, K. J., Shultz, A. J., Title, P. O., Mason, N. A., Barker, F. K., Klicka, J., \ldots{} Lovette, I. J. (2014). Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds. \emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.

\leavevmode\hypertarget{ref-Chamberlain2018}{}%
Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}

\leavevmode\hypertarget{ref-Chamberlain2013}{}%
Chamberlain, S. A., \& Szöcs, E. (2013). taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}. \emph{F1000Research}, \emph{2}(191), 1--29. \url{https://doi.org/10.12688/f1000research.2-191.v2}

\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
Chaves, J. A., Hidalgo, J. R., \& Klicka, J. (2013). Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini). \emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.

\leavevmode\hypertarget{ref-claramunt2015new}{}%
Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.

\leavevmode\hypertarget{ref-Criscuolo2006}{}%
Criscuolo, A., Berry, V., Douzery, E. J., \& Gascuel, O. (2006). SDM: A fast distance-based approach for (super)tree building in phylogenomics. \emph{Systematic Biology}, \emph{55}(5), 740--755. \url{https://doi.org/10.1080/10635150600969872}

\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E. J. (2018). A phylogenomic framework and timescale for comparative studies of tunicates. \emph{BMC Biology}, \emph{16}(1), 1--14.

\leavevmode\hypertarget{ref-Eastman2013}{}%
Eastman, J. M., Harmon, L. J., \& Tank, D. C. (2013). Congruification: Support for time scaling large phylogenetic trees. \emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691. \url{https://doi.org/10.1111/2041-210X.12051}

\leavevmode\hypertarget{ref-edgar2004muscle}{}%
Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.

\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}

\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
GBIF Secretariat. (2022). GBIF Backbone Taxonomy. \emph{Checklist dataset}, (Online Resource accessed via GBIF.org). Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}

>>>>

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

DateLife's core software application consists of the R package \texttt{datelife}. Its current stable version -- v0.6.8, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Pennell et al., 2014),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.

@@

DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.

DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string. Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.

@@

A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database. Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
These matching pruned chronograms are referred to as source chronograms.
Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

@@

DateLife requires a tree topology to summarize age data upon.
We recommend that users provide as input a tree topology from the literature, or one of their own making. If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
If source chronograms do not form a grove, the supertree reconstruction will fail.

@@

For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).

@@

Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).

@@

\textbf{\emph{Summarizing search results.--}}
DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).

@@

\textbf{\emph{Creating a query.--}}
To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
which gets all recognized species names within a named group from a taxonomy of choice.
Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group. Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.

@@

\textbf{\emph{Summarizing search results.--}}
The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).

@@

We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).

@@

In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.

Making up branch lengths is a common practice in scientific publications: Jetz et al. (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
Alström, P., Hooper, D. M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T. D. (2014). Discovery of a relict lineage and monotypic family of passerine birds. \emph{Biology Letters}, \emph{10}(3), 20131067.

\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.

\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
Antonelli, A., Hettling, H., Condamine, F. L., Vos, K., Nilsson, R. H., Sanderson, M. J., \ldots{} Vos, R. A. (2017). {Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}. \emph{Systematic Biology}, \emph{66}(2), 153--166. \url{https://doi.org/10.1093/sysbio/syw066}

\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
Archie, J., Day, W. H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F. J., \& Swofford, D. (1986). {The Newick tree format}. Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}

\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
Avibase. (2022). {Yellow-throated Bunting}. \emph{{Avibase - The World Bird Database}}, (Online Resource). Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}

\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
Bapst, D. W. (2012). {Paleotree: An R package for paleontological and phylogenetic analyses of evolution}. \emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807. \url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}

\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
Barba-Montoya, J., Reis, M. dos, Schneider, H., Donoghue, P. C., \& Yang, Z. (2018). Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution. \emph{New Phytologist}, \emph{218}(2), 819--834.

\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.

\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
Barker, F. K., Burns, K. J., Klicka, J., Lanyon, S. M., \& Lovette, I. J. (2013). Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds. \emph{{Systematic Biology}}, \emph{62}(2), 298--320.

\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
Barker, F. K., Burns, K. J., Klicka, J., Lanyon, S. M., \& Lovette, I. J. (2015). New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. \emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.

\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
Barker, F. K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J. (2004). Phylogeny and diversification of the largest avian radiation. \emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.

\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T. (2005). African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.

\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J. A., Mozzherin, D., Rees, T., \ldots{} Enquist, B. J. (2013). {The taxonomic name resolution service: An online tool for automated standardization of plant names}. \emph{{BMC Bioinformatics}}, \emph{14}(1). \url{https://doi.org/10.1186/1471-2105-14-16}

\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
Britton, T., Anderson, C. L., Jacquet, D., Lundqvist, S., \& Bremer, K. (2007). {Estimating Divergence Times in Large Phylogenetic Trees}. \emph{{Systematic Biology}}, \emph{56}(788777878), 741--752. \url{https://doi.org/10.1080/10635150701613783}

\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
Bryson Jr, R. W., Chaves, J., Smith, B. T., Miller, M. J., Winker, K., Pérez-Emán, J. L., \& Klicka, J. (2014). Diversification across the new world within the `blue'cardinalids (aves: cardinalidae). \emph{Journal of Biogeography}, \emph{41}(3), 587--599.

\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.

\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
Burns, K. J., Shultz, A. J., Title, P. O., Mason, N. A., Barker, F. K., Klicka, J., \ldots{} Lovette, I. J. (2014). Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds. \emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.

\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
Chamberlain, S. A., \& Szöcs, E. (2013). {taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}. \emph{{F1000Research}}, \emph{2}(191), 1--29. \url{https://doi.org/10.12688/f1000research.2-191.v2}

\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
Chamberlain, S. A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G. (2019). \emph{{taxize: Taxonomic information from around the web}}. Retrieved from \url{https://github.com/ropensci/taxize}

\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
Chaves, J. A., Hidalgo, J. R., \& Klicka, J. (2013). Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini). \emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.

\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.

\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
Criscuolo, A., Berry, V., Douzery, E. J. P., \& Gascuel, O. (2006). {SDM: A fast distance-based approach for (super)tree building in phylogenomics}. \emph{{Systematic Biology}}, \emph{55}(5), 740--755. \url{https://doi.org/10.1080/10635150600969872}

\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E. J. (2018). A phylogenomic framework and timescale for comparative studies of tunicates. \emph{BMC Biology}, \emph{16}(1), 1--14.

\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
Eastman, J. M., Harmon, L. J., \& Tank, D. C. (2013). {Congruification: Support for time scaling large phylogenetic trees}. \emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691. \url{https://doi.org/10.1111/2041-210X.12051}

\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
